The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks

### 2024-11-27 paper.Rmd
updated text in sim 1

As shown in Figure \ref{fig:dimerization}, both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (@MartinezFernandez2022; @Mouret2006). The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction.

@@

As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers. Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates based on the monomers involved as well as the level of UV radiation (@Mouret2006).

#### Implications and Real-World Applications
Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer.The Gillespie  algorithm allows us to realistically model these reactions, as it allows us to demonstrate that our parameters of interest change over time, rather than following a general logarithmic curve.This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels. 
Using the Gillespie algorithm to model molecular reactions, such as the dimerization of thymine shown in this simulation, has wide-ranging implications in fields such as genetics, cancer research, pharmacology, environmental science, and biotechnology. By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems, enabling better prediction of outcomes in disease development, therapy, and environmental impact studies.

>>>>

As shown in Figure \ref{fig:dimerization}, both CPD and 64-PP formation follow logarithmic growth, which aligns with expectations based on specific UV radiation conditions and previously reported quantum yields (@MartinezFernandez2022; @Mouret2006). The number of reactions per unit time levels off as pyrimidine pairs decrease, which is consistent with the propensity functions, as the propensity of 64-PP formation is much significantly than the propensity of CPD formation (@Mouret2006). This plateau suggests that, as the number of available reactants decreases, the rate of reaction slows down.
Due to the stochastic nature of molecular reactions, this simulation demonstrates how the Gillespie algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each individual reaction event. These fluctuations are especially important in low-molecule systems, where randomness significantly impacts the overall reaction dynamics, providing a more accurate representation of the molecular behavior compared to deterministic models.

@@

As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers. Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates dependent on the monomers involved as well as the level of UV radiation (@Mouret2006).

#### Implications and Real-World Applications
The Gillespie algorithm offers profound implications for understanding molecular reactions, particularly in systems where stochastic effects are prominent, such as the dimerization. This process, highlighted in Figure \ref{fig:dimerization}, is widely studied, as it can lead to mutations potentially resulting in skin cancer, when not repaired. The Gillespie  algorithm allows us to realistically model these reactions, as it demonstrates the probabilistic nature of DNA damage, capturing fluctuations in molecular interactions that deterministic models cannot, providing a deeper understanding of molecular events, insights into mutation rates, and the impact of environmental factors like UV radiation on genetic integrity. The algorithm's ability to model these stochastic processes is crucial in areas such as genetics, where random molecular events can significantly influence disease outcomes. 
This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels. Using the Gillespie algorithm to model molecular reactions has wide-ranging implications in fields such as genetics, cancer research, pharmacology, environmental science, and biotechnology. By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems, enabling better prediction of outcomes, such as in disease development, medicine, and environmental impact studies.
